---
title: "Model Selection --- Data lab"
author: "Dmitry Kondrashov & Stefano Allesina"
date: "Fundamentals of Biological Data Analysis -- BIOS 26318"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
urlcolor: blue
---

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  comment   = "#",
  warning = FALSE,
  message = FALSE,
  #results   = "asis",
  # collapse  = TRUE,
  fig.align = "center")
```

```{r, message = FALSE}
library(tidyverse) # our friend the tidyverse
library(readxl) # to read excel files
source("general_code/read_xls_from_url.R") # function  to read excel from URL
```

## Litter size

[Data](https://doi.org/10.5061/dryad.g4q84) from the paper:

> Kallio ER, Helle H, Koskela E, Mappes T, Vapalahti O (2015) [Age-related effects of chronic hantavirus infection on female host fecundity](https://doi.org/10.1111/1365-2656.12387). Journal of Animal Ecology 84(5): 1264–1272. 

```{r}
# read the data
# use short URL
# original: datadryad.org/bitstream/handle/10255/dryad.84805/Litter_data.xlsx
dat <- read_xlsx_from_url("https://tinyurl.com/y88w7xe6")
# show the data
head(dat)
# plot litter sizes
dat %>% ggplot(aes(x = litter_size)) + geom_histogram(binwidth = 1)
```

### Poisson

Let's write the likelihood function for the Poisson distribution
```{r}
log_likelihood_Poisson <- function(my_data, my_lambda){
  return(sum(log(dpois(my_data, my_lambda))))
}
```

For example:
```{r}
log_likelihood_Poisson(dat$litter_size, 3)
log_likelihood_Poisson(dat$litter_size, 5)
```

But what is the maximum likelihood estimate for $\lambda$?

```{r}
# use optim(initial guess, function to maximize, other named parameters)
# we use method Brent that looks for the maximum in an interval
# we set control = list(fnscale = -1) 
# to maximize instead of minimize (default) the function
find_lambda <- optim(5, log_likelihood_Poisson, 
                     my_data = dat$litter_size, 
                     method = "Brent", lower = 0, upper = 10,
                     control = list(fnscale = -1))
find_lambda
```

Let's store the best parameter and maximum likelihood:
```{r}
lambda_hat <- find_lambda$par
max_likelihood_Poisson <- find_lambda$value
```

Using Poisson regression

```{r}
summary(glm(dat$litter_size ~ 1., family = "poisson"))
exp(1.67730)
```

Cleaning messy output with `broom`
```{r}
library(broom)
model1 <- glm(litter_size ~ 1., family = "poisson", data = dat)
model1_fit <- tidy(model1)
```
Compute AIC:
```{r}
AIC(model1)
```

Now let's see whether the infection status can help us explain the litter size:

```{r}
model2 <- glm(litter_size ~ PUUV_INF, family = "poisson", data = dat)
summary(model2)
AIC(model2)
```

Not really! In fact, the effect is to simply add a parameter.

Now let's look at over/underdispersion:
```{r}
summary(glm(litter_size ~ 1., family = "quasipoisson", data = dat))
```
The data is highly underdispersed! The variance is about 1/4 of what expected using a Poisson distribution. Hence, using a more flexible distribution should help considerably:

```{r}
model3 <- MASS::glm.nb(litter_size ~ 1., data = dat)
summary(model3)
AIC(model3)
```

