---
title: "Distributions and their properties: Data Lab"
author: "Dmitry Kondrashov & Stefano Allesina"
date: "Fundamentals of Biological Data Analysis -- BIOS 26318"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
urlcolor: blue
---

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  comment   = "#",
  #results   = "asis",
  # collapse  = TRUE,
  fig.align = "center")
```

```{r, message = FALSE}
library(MASS) # to fit distributions
library(tidyverse) # our friend the tidyverse
library(readxl) # to read excel files
source("../general_code/read_xls_from_url.R") # function  to read excel from URL
```

# Visualizing 1-dimensional distributions
We start by analyzing a dataset reporting the incubation period of a novel type of influenza A of avian origin that emerged in China in 2013. The data are taken from:

> Virlogeux V, Yang J, Fang VJ, Feng L, Tsang TK, Jiang H, Wu P, Zheng J, Lau EHY, Qin Y, Peng Z, Peiris JSM, Yu H, Cowling BJ (2016) [Association between the severity of influenza A(H7N9) virus infections and length of the incubation period.](https://doi.org/10.1371/journal.pone.0148506) PLOS ONE 11(2): e0148506. 

We can read the data in directly from the internet:
```{r}
# Original URL
# datadryad.org/bitstream/handle/10255/dryad.104362/data_h7n9_severity.csv
dat <- read.csv("https://tinyurl.com/y8yfnq5t")
head(dat)
```
where `Case.No.` is an identifier of the record, `IncP_min` and `IncP_max` are the minimum (maximum) incubation period, `age` is the age of the patient, `sex_status` is `0` for female and `1` for male, and `death_status` reports whether the patient died. 

Because we're going to work with only a few columns, let's select drop the rest:
```{r}
# we will see this in more details later in the course
# as the name of the command implies, we're selecting certain cols
dat <- select(dat, IncP_max, age, sex_status, death_status)
```

## Histograms
A simple way to visualize a distribution is to plot a histogram: data are binned, and the height of the bin represents counts (or frequencies). 
```{r, message=FALSE, warning=FALSE}
ggplot(data = dat) + aes(x = age) + geom_histogram()
# you can control the width of the bins
ggplot(data = dat) + aes(x = age) + geom_histogram(binwidth = 4)
# or set the desired number of bins
ggplot(data = dat) + aes(x = age) + geom_histogram(bins = 40)
# you can control the color of the boxes (both border and fill)
ggplot(data = dat) + aes(x = age) + 
  geom_histogram(bins = 40, fill = "lightblue", color = "darkred")
```
In many cases, a density plot (i.e., where the bins have been "interpolated") can be used instead:
```{r}
ggplot(data = dat) + aes(x = age) + geom_density()
```
In this case, we can see clearly that the distribution is bimodal (i.e., has two peaks).

You can also combine both the histogram and the density calling:
```{r}
ggplot(data = dat) + aes(x = age) + 
  geom_histogram(bins = 40,  aes(y = ..density..)) + 
  geom_density()
```

where the command `aes(y = ..density..)` transforms the height of the bars from counts to densities. 

## Adding features to the histogram
Suppose we want to show where the mean (median, etc.) of the distribution is. We can combine the histogram with a vertical line:

```{r, message=FALSE, warning=FALSE}
ggplot(data = dat) + aes(x = age) +
  geom_histogram() + 
  geom_vline(aes(xintercept = mean(age)), linetype = 2) +
  geom_vline(aes(xintercept = median(age)), linetype = 3, color = "red")
```

# Contrasting 1-dimensional distributions
We can plot more than one histogram on the same plot. For example, let's see whether the distribution of ages differs between men and women:
```{r, message=FALSE, warning=FALSE}
ggplot(data = dat) + 
  aes(x = age, fill = as.factor(sex_status)) + 
  geom_histogram(position = "identity", alpha = 0.4)
```

where we colored the bars by sex (`as.factor` transfoms the numbers `0` and `1` into "categories"), and we set the transparency (`alpha = 0.4`) to be able to see both distributions. When we set the position to `identity` the bars for the two histogram are located at the same positions. To arrange them side-to-side we can use `dodge`:

```{r}
ggplot(data = dat) + 
  aes(x = age, fill = as.factor(sex_status)) + 
  geom_histogram(position = "dodge", alpha = 0.4)
```

Another way to contrast distributions is by using box-, dot- or violin-plots. In this case, we look at the distribution "from above", and concentrate on the location of the median, and the width of the distribution. These plots allow to quickly assess whether the distributions differ:
```{r}
# boxplot
ggplot(data = dat) + 
  aes(x = as.factor(sex_status), y = age) + 
  geom_boxplot()
# dotplot
ggplot(data = dat) + 
  aes(x = as.factor(sex_status), y = age) + 
  geom_dotplot(stackdir = "center", binaxis = "y", dotsize = 0.5)
# violin plot
ggplot(data = dat) + 
  aes(x = as.factor(sex_status), y = age) + 
  geom_violin()
```

# Visualizing 2-dimensional distributions
Now let's look at the incubation period vs. age
```{r}
ggplot(data = dat) + aes(x = IncP_max, y = age) + geom_bin2d() + scale_x_sqrt()
```

What we've produced is a 2D histogram. Maybe a density plot would be more informative:

```{r}
ggplot(data = dat) + aes(x = IncP_max, y = age) + geom_density_2d() + scale_x_sqrt()
```


# Types of distributions in biological data

Whenever you want to model your data using a statistical distribution, you need to think about the characteristics of your data. 

## Discrete vs. continuous
Discrete means that the quantity can only assume certain values (e.g., number of births can only be integrer, age in years is integer, male/female is categorical, taking only two values, etc.). Continuous means that you value can fall in between any two values (e.g., height, age when measured continuously, weight, solar radiance, temperature, etc.).

To see the difference, we're going to look at number of offspring and weight of the mother using the data from:

> O'Dea RE, Vega-Trejo R, Head ML, Jennions MD (2015) [Maternal effects on offspring size and number in mosquitofish](https://doi.org/10.1002/ece3.1577), Gambusia holbrooki. Ecology and Evolution 5(14): 2945-2955. 

```{r}
# Original URL
#https://datadryad.org/bitstream/handle/10255/dryad.90634/DryadData.xlsx
dat <- read_xlsx_from_url("https://tinyurl.com/y9v8lmks")
dat
```

You can see that the `Number_of_Offspring` is discrete, while the `Female_Size` is continuous.

For example:
```{r}
table(dat$Number_of_Offspring)
ggplot(data = dat) + aes(x = Female_Size) + geom_histogram(binwidth = 2)
ggplot(data = dat) + aes(x = Offspring_Size) + geom_histogram(binwidth = 0.05)
```

Note that in some cases, it is convenient to treat discrete variables as continuous (e.g., when you have very many possible values, such as number of individual ants across several nests).

## Support
The second thing you need to determine is whether the variable can take values in the whole number line (i.e., unbounded), or rather has some bounds. For example, the number of offspring, the weight and height cannot be negative; a probability can only assume values between 0 and 1; the number of heads in $k$ flips of a coin is bounded by $0$ and $k$; etc.

Again, in some cases you might want to treat as unbounded quantities that have natural bounds: for example, a Normal distribution well describes the height of the people in this room, though clearly no one can be -2.5m tall (or even 5m tall).

## Multimodality
When we plot the histogram of the variable, we can observe a single "peak" (unimodal) or more than one (multi-modal). Often, observing two peaks means that we are observing a mixture of two distributions (e.g., the influenza case above). In such cases, we want to try and find what the origin of the multi-modality seems to be.

## Skewness
Finally, you want to check whether your variable shows a distribution that is symmetric about the mean, or rather has a thicker upper tail than lower tail (positive skew) or vice versa (negative skew). 


# Do the data follow a certain distribution?
Suppose we want to check whether our data is well matched by a certain distribution. For example, let's produce data that should follow a normal distribution with mean 3 and standard deviation 1.5:

```{r}
set.seed(101)
test_data <- tibble(x = rnorm(n = 500, mean = 3, sd = 1.5))
```

We can find the best-fitting parameters for our data and a distribution of choice:
```{r}
# example: find best-fitting Normal
my_normal <- fitdistr(test_data$x, densfun = "normal")
# note the slight discrepancies
my_normal
```

And produce a so-called Q-Q plot (quantile-quantile plot): for each quantile of our data, the theoretical value is computed for the best-fitting "theoretical" distribution. If the data were to follow the distribution closely, you should find all the points lying on the 1:1 line:
```{r}
ggplot(test_data, aes(sample = x)) +
  stat_qq(distribution = qnorm, dparams = my_normal$estimate) +
  stat_qq_line(distribution = qnorm, dparams = my_normal$estimate) + 
  geom_abline(intercept = 0, slope = 1, linetype = 2, col = "red") +
  ggtitle("Q-Q plot assuming best-fitting Normal distribution")
```

## But what if we choose the wrong distribution?
Suppose that we choose the values from a Uniform distribution:
```{r}
test_data <- tibble(x = runif(n = 500))
```

And that we fit an exponential distribution:
```{r}
my_exponential <- fitdistr(test_data$x, densfun = "exponential")
ggplot(test_data, aes(sample = x)) +
  stat_qq(distribution = qexp, dparams = my_exponential$estimate) +
  stat_qq_line(distribution = qexp, dparams = my_exponential$estimate) + 
  ggtitle("Q-Q plot assuming best-fitting Exponential distribution")
```

Or a Weibull distribution:
```{r}
my_weibull <- fitdistr(test_data$x, densfun = "weibull")
ggplot(test_data, aes(sample = x)) +
  stat_qq(distribution = qweibull, dparams = my_weibull$estimate) +
  stat_qq_line(distribution = qweibull, dparams = my_weibull$estimate) + 
  ggtitle("Q-Q plot assuming best-fitting Weibull distribution")
```

On the other hand, if we use a distribution that can be made into a Uniform distribution by setting parameters to particular values (e.g., $\Beta(1,1)$ is equal to Uniform between 0 and 1), we find a good agreement:
```{r}
# for some distributions, you need to provide a starting point 
# for the search of parameters
my_beta <- fitdistr(test_data$x, densfun = "beta", 
                    start = list(shape1 = 1.5, shape2 = 2.1))
ggplot(test_data, aes(sample = x)) +
  stat_qq(distribution = qbeta, dparams = my_beta$estimate) +
  stat_qq_line(distribution = qbeta, dparams = my_beta$estimate) + 
  ggtitle("Q-Q plot assuming best-fitting Beta distribution")
```

```{r}
# and in fact the best-fitting Beta is close to Uniform
my_beta
```


