---
title: "Clustering methods"
author: "Dmitry Kondrashov & Stefano Allesina"
date: "Fundamentals of Biological Data Analysis -- BIOS 26318"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
urlcolor: blue
---

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  comment   = "#",
  warning = FALSE,
  message = FALSE,
  #results   = "asis",
  # collapse  = TRUE,
  fig.align = "center")
```

## Goals

  * Learn about partitional clustering
  * Learn about hierarchical clustering
  * Use clustering validation methods
  * Apply different methods to larger data sets

```{r}
library(tidyverse) # our friend the tidyverse
library(ggfortify) 
library(factoextra) 
library(NbClust)
library(fpc)
library(clustertend)
```


## K-means algorithm

 * divide data into K clusters
 * calculate centroids for each
 * go through each data point until nothing changes
   + calculate distance to each centroid
   + assign to nearest centroid
   + recalculate centroids for the two affected clusters

```{r}
#set.seed(20)
iris.scaled <- scale(iris[, -5])
iris_km <- kmeans(iris.scaled, 3)
iris_km
table(iris_km$cluster, iris$Species)

fviz_cluster(list(data = iris.scaled, cluster = iris_km$cluster),
ellipse.type = "norm", geom = "point", stand = FALSE, palette = "jco", ggtheme = theme_classic())
```


### Assumptions of K-means algorithm

  * There is a meaningful distance measure
  * Clusters are roughly spherical
  * Clusters are of similar size


```{r}
#Generate random data which will be first cluster
clust1 <- data_frame(x = rnorm(200), y = rnorm(200))
#Generate the second cluster which will ‘surround’ the first cluster
clust2 <- data_frame(r = rnorm(200, 15, .5), theta = runif(200, 0, 2 * pi),
                 x = r * cos(theta), y = r * sin(theta)) %>%
  dplyr::select(x, y)
#Combine the data
dataset_cir <- rbind(clust1, clust2)
#see the plot
dataset_cir %>% ggplot() + aes(x=x, y=y) + geom_point()
```



```{r}
#Fit the k-means model
k_clust_spher1 <- kmeans(dataset_cir, centers=2)
#Plot the data and clusters
fviz_cluster(list(data = dataset_cir, cluster = k_clust_spher1$cluster),
ellipse.type = "norm", geom = "point", stand = FALSE, palette = "jco", ggtheme = theme_classic())

```

```{r}
#Make the first cluster with 1000 random values
clust1 <- data_frame(x = rnorm(1000), y = rnorm(1000))
#Keep 10 values together to make the second cluster
clust2 <- data_frame(x=c(5,5.1,5.2,5.3,5.4,5,5.1,5.2,5.3,5.4),y=c(5,5,5,5,5,4.9,4.9,4.9,4.9,4.9))
#Combine the data
dataset_uneven <- rbind(clust1,clust2)
dataset_uneven %>% ggplot() + aes(x=x, y=y) + geom_point()
```
```{r}
k_clust_spher3 <- kmeans(dataset_uneven, centers=2)
fviz_cluster(list(data = dataset_uneven, cluster = k_clust_spher3$cluster),
ellipse.type = "norm", geom = "point", stand = FALSE, palette = "jco", ggtheme = theme_classic())
```



### Exercise: PCA sommelier

The file `Wine.csv` contains several measures made on 178 wines from Piedmont, produced using three different grapes (column `Grape`, with 1 = Barolo, 2 = Grignolino, 3 = Barbera). Use the 13 measured variables (i.e., all but `Grape`) to perform a PCA, and use k-means to cluster the data in PC space. Can you recover the right classification of grapes?

```{r}
wine <- read_csv("data/Wine.csv")
# make into a matrix for PCA
mat <- wine %>% select(-Grape) %>% as.matrix()
# perform PCA by scaling and centering all 13 variables
pca <- prcomp(mat, center = TRUE, scale. = TRUE)
# pca$x contains the coordinates
# let's see how are we doing
ggplot(cbind(wine, as_tibble(pca$x))) + 
  aes(x = PC1, y = PC2, colour = factor(Grape)) + 
  geom_point()
```

Now let's apply k-means to divide the data into three clusters

```{r}
cl <- kmeans(pca$x, centers = 3)
ggplot(wine %>% add_column(cluster = cl$cluster)) + 
  aes(x = Grape, y = cluster, colour = factor(Grape)) + 
  geom_jitter()
```

You can see that we correctly classify all the wines for grapes 1 and 3, while some of the ones from grape 2 are misclassified. 

```{r}
# find convex hull for each cluster
# i.e. minimal convex polygon containing all points
hull <- pca$x %>% as_tibble() %>% 
  select(PC1, PC2) %>% 
  add_column(cluster = cl$cluster) %>% 
  group_by(cluster) %>% 
  slice(chull(PC1, PC2))

# plot the convex hulls
pl <- hull %>% ggplot(aes(x = PC1, y = PC2, group = cluster, fill = factor(cluster))) + geom_polygon(alpha = 0.5)
  
show(pl)

# now add the points
pl + geom_point(data = pca$x %>% as_tibble() %>% 
         select(PC1, PC2) %>% 
         add_column(Grape = wine$Grape, cluster = cl$cluster), 
  aes(x = PC1, y = PC2, colour = factor(Grape), shape = factor(Grape)))

table(cl$cluster, wine$Grape)
```

You can see that we are misclassifying the grapes that are at the border of the cluster or nestled among the points belonging to a different grape. But the outcome is quite good: we would have misclassified only 6 wines out of 178 (3.3%).

## Hierarchical clustering


### Agglomerative clustering

Start with single data points as "clusters," then iteratively combine the closest pair of clusters. The closenest may be defined in the following ways:

 1. Single Linkage: In single linkage, we define the distance between two clusters as the minimum distance between any single data point in the first cluster and any single data point in the second cluster. 

 2. Complete Linkage: In complete linkage, we define the distance between two clusters to be the maximum distance between any single data point in the first cluster and any single data point in the second cluster.

 3. Average Linkage: In average linkage, we define the distance between two clusters to be the average distance between data points in the first cluster and data points in the second cluster. 

4. Centroid Method: In centroid method, the distance between two clusters is the distance between the two mean vectors of the clusters.  

 5. Ward’s Method: This method does not directly define a measure of distance between two points or clusters. It is an ANOVA based approach. One-way univariate ANOVAs are done for each variable with groups defined by the clusters at that stage of the process.  At each stage, two clusters merge that provide the smallest increase in the combined error sum of squares. 


```{r}
# Use hcut() which compute hclust and cut the tree
cir_hc <- hcut(dataset_cir, k = 2, hc_method = "single")
# Visualize dendrogram
fviz_dend(cir_hc, show_labels = FALSE, rect = TRUE)
# Visualize cluster
fviz_cluster(cir_hc, ellipse.type = "convex")
```

```{r}
# Use hcut() which compute hclust and cut the tree
uneven_hc <- hcut(dataset_uneven, k = 2, hc_method = "single")
# Visualize dendrogram
#fviz_dend(uneven_hc, show_labels = FALSE, rect = TRUE)
# Visualize cluster
fviz_cluster(uneven_hc, ellipse.type = "convex")
```

### Cluster the irises using hierarchical methods
Try different methods and see which one generates the best results

```{r}
# Hierarchical clustering
# ++++++++++++++++++++++++
# Use hcut() which compute hclust and cut the tree
iris_hc <- hcut(iris.scaled, k = 3, hc_method = "single")
# Visualize dendrogram
fviz_dend(iris_hc, show_labels = FALSE, rect = TRUE)
# Visualize cluster
fviz_cluster(iris_hc)
table(iris_hc$cluster, iris$Species)
```


### Taste the wine again!

For the wine data set, use hierarchical clustering on the PCA - reduced data set and compare the results to the complete data set. Which hierarchical clustering method performs the best?

```{r}

```




## Clustering analysis and validation

### Hopkins statistic
Comparing the mean nearest-neighbor distance between uniformly generated sample points and mean nearest-neighbor distance within the data set. 
$$
H = 1 - \frac{\sum u^d_i}{\sum u^d_i + \sum w^d_i}
$$
This quantifies the "clustering tendency" of the data set.
```{r}
# Check Cluster Tendency--Hopkins Statistic
hopkins(iris.scaled, n = 30) # n should be about 20% of the data
# run a couple times to sample repeatedly
```

If H is below 0.5 reject the null hypothesis, which is that the data are generated by a Poisson point process (uniformly distributed.)

```{r}
# Visual Assessment of Cluster Tendency
fviz_dist(dist(iris.scaled), show_labels = FALSE)+ labs(title = "Iris Data")
```
- Red is high similarity (low dissimilarity)
- Blue is low similarity (high dissimilarity)

### Elbow method

```{r}
# Elbow method
fviz_nbclust(iris.scaled, kmeans, method = "wss") + geom_vline(xintercept = 2, linetype = 2)+
labs(subtitle = "Elbow method for K-means")
```
### Silhouette Plot
Measures how similar an object $i$ is to the other objects in its same cluster versus the objects outside of its cluster; $S_i$ values range from -1 to 1. Close to 1 means very similar to objects in its own group and dissimilar to others
```{r}
# Silhouette method
fviz_nbclust(iris.scaled, kmeans, method = "silhouette")+ labs(subtitle = "Silhouette method for k-means")
```

### Lazy way: use all the methods!

```{r}
nb <- NbClust(iris.scaled, distance = "euclidean", min.nc = 2,
        max.nc = 10, method = "kmeans")
fviz_nbclust(nb)
```



### Validation

One common approach to validating clustering is to use the approach called bootstrappingm which involves repeatedly sampling from the data set, running the clustering algorithm and comparing the results. One algorithm uses the Jaccard coefficient to quantify similarity between sets, which is defined as the number of points in the intersection of the two sets (those which are in both sets), divided by the number of points in the union of the two sets (the point that are in either one or the other set):

$$
J = \frac{ \vert A \cap B \vert }{\vert A \cup B \vert}
$$
The vertical lines indicate the number of points (cardinality) in the set.


```{r}
k <- 3
cboot.hclust <- clusterboot(iris.scaled, clustermethod=kmeansCBI, k= k)
print(cboot.hclust)

#cboot.hclust <- clusterboot(bcdata, clustermethod=hclustCBI,
       #                    method="single", k=2)
```


## Application to breast cancer data

```{r}
# Import Breast Cancer Data Set
fulldata <- read_csv("data/Wisconsin_Breast_Cancers.csv")
bcdata <- fulldata[,2:10]
class <- array(data= fulldata[,11])
head(fulldata)
```

```{r}
# Visually Inspect Data (PCA)
fviz_pca_ind(prcomp(bcdata), title = "PCA - Breast Cancer data", geom = "point", ggtheme = theme_classic())
```

```{r}
bc_km <- kmeans(scale(bcdata), 2)
bc_km
table(bc_km$cluster, fulldata$Class)
#irisCluster$cluster <- as.factor(irisCluster$cluster)
#ggplot(iris, aes(Petal.Length, Petal.Width, color = iris$cluster)) + geom_point()

fviz_cluster(list(data = bcdata, cluster = bc_km$cluster),
ellipse.type = "norm", geom = "point", stand = FALSE, palette = "jco", ggtheme = theme_classic())
```

```{r}
# Use hcut() which compute hclust and cut the tree
bc_hc <- hcut(scale(bcdata), k = 2, hc_method = "ward")
# Visualize dendrogram
fviz_dend(bc_hc, show_labels = FALSE, rect = TRUE)
# Visualize cluster
fviz_cluster(bc_hc, ellipse.type = "convex")
table(bc_hc$cluster, fulldata$Class)
```

## References:

1. https://www.r-bloggers.com/exploring-assumptions-of-k-means-clustering-using-r/
2. https://onlinecourses.science.psu.edu/stat505/node/143/
3. https://github.com/hhundiwala/hierarchical-clustering
4. https://www.r-bloggers.com/bootstrap-evaluation-of-clusters/
